---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

First, we load the dataset and transform variables to fit the analysis.
```{r}
echo=TRUE

unzip("activity.zip")
data<-read.csv("activity.csv",header=TRUE)
data$date<-as.Date(data$date,"%Y-%m-%d")
```

## What is mean total number of steps taken per day?
Missing values will be omitted for this part of the analysis.

First, we create a histogram of total steps taken each day.
```{r}
total.steps <- aggregate(data$steps, list(Date=data$date), FUN="sum")
colnames(total.steps) <- c("date", "steps")
total.steps.1<-na.omit(total.steps)

library(ggplot2)
ggplot(data=total.steps.1,aes(date,steps))+
  geom_bar(stat="identity")+
  labs(title="Total Number of Steps",x="Date")
```

Then we calculate the mean and median to total steps taken each day.
```{r}
mean.total.steps<-mean(total.steps.1$steps)
median.total.steps<-median(total.steps.1$steps)
```

## What is the average daily activity pattern?
First, we create a histogram of average steps taken for all 5-minute intervals across a day.
```{r}
data.1<-na.omit(data)
pattern<-aggregate(list(steps=data.1$steps),list(interval=data.1$interval),FUN="mean")

ggplot(data=pattern,aes(x=interval,y=steps))+
  geom_line()+
  labs(title="Average Daily Activity Pattern",x="interval",y="average steps")

```

Then we identify the 5-minute interval that on average has maximum number of steps.
```{r}
pattern[pattern$steps==max(pattern$steps),]
```


## Imputing missing values
First, we identify the total number of rows that contain missing values.
```{r}
sum(is.na(data$steps))
```

Next, we replace the missing values with the mean steps taken in the corresponding 5-minute intervals
```{r}
data.filled<-data

for(i in 1:nrow(data.filled)){
  if(is.na(data.filled$steps[i])){
    interval<-data.filled$interval[i]
    filled.value<-pattern$steps[pattern$interval==interval]
    data.filled$steps[i]<-filled.value}
}

```

Then we create a histrogram based on the new dataset without missing values.
```{r}
total.steps.filled <- aggregate(data.filled$steps, list(Date=data.filled$date), FUN="sum")
colnames(total.steps.filled) <- c("date", "steps")

ggplot(data=total.steps.filled,aes(date,steps))+
  geom_bar(stat="identity")+
  labs(title="Total Number of Steps")
```

For the new dataset, the mean and median total steps taken each day are calculated as follows.
```{r}
mean.total.steps.filled<-mean(total.steps.filled$steps)
median.total.steps.filled<-median(total.steps.filled$steps)

mean.total.steps.filled-mean.total.steps #compare the means
median.total.steps.filled-median.total.steps #compare the medians
```

## Are there differences in activity patterns between weekdays and weekends?
First, we add a new column that indicate weekday status to the dataset with no missing values.
```{r}
weekdays<-NULL

for(i in 1:nrow(data.filled)){
  day<-weekdays(data.filled$date[i])
  if (day %in% c("Saturday","Sunday")){weekdays<-c(weekdays,"weekend")}
  else {weekdays<-c(weekdays,"weekday")}
}

data.filled<-cbind(data.filled,weekdays)

```

Next, we creatte a time series plot that shows steps taken across 5-minute intervals on weekdays vs. weekends.
```{r}
pattern.weekdays<-aggregate(list(steps=data.filled$steps),list(interval=data.filled$interval,weekdays=data.filled$weekdays),FUN="mean")

library(lattice)
xyplot(pattern.weekdays$steps ~ pattern.weekdays$interval | pattern.weekdays$weekdays, 
       type = "l", layout = c(1, 2), xlab = "5-min Interval", ylab = "Average number of steps")
```

